\documentclass[11pt]{article}

% --- Packages (The "plugins" of LaTeX) ---
\usepackage[utf8]{inputenc} % Character encoding
\usepackage{geometry}       % Page margins
 \geometry{margin=1in}
\usepackage{graphicx}       % For inserting images
\usepackage{amsmath}        % For math equations
\usepackage{listings}
\usepackage{xcolor} % Required for syntax highlighting colors

% Define colors for Python syntax
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
\definecolor{termback}{rgb}{0.1,0.1,0.1}
\definecolor{termtext}{rgb}{0.9,0.9,0.9}
\definecolor{lightgray}{rgb}{0.9,0.9,0.9}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstdefinestyle{terminal}{
    backgroundcolor=\color{white},   % White background
    basicstyle=\ttfamily\small\color{black}, % Black text
    frame=single,                    % Simple border
    rulecolor=\color{lightgray},     % Light gray border color
    breaklines=true,                 % Wrap long lines
    numbers=none,                    % No line numbers for output
    captionpos=b,                    % Caption at the bottom
    xleftmargin=10pt,                
    xrightmargin=10pt,
    framesep=5pt
}

\lstset{style=mystyle}

% --- Document Information ---
\title{DE 414: Practical 2 Report}
\author{Amy McDermott (26911264)}
\date{\today}

\begin{document}

\maketitle

\section*{Introduction}
This report documents the implementation of a linear regression model using Python. The model is trained on the MNIST dataset with the aim to 
correctly classify digits. This report includes the code used for training the model, as well as the results obtained.

\section{Question 1: Load training and test data}
The MNIST dataset is loaded from a numpy archive file. This functionality is implemented in the function \texttt{read\_mnist\_npz()}, which is then called to load the training and test data.
The file name for the numpy archive is passed into this function. The shapes of the training and test data are then printed to the terminal to confirm that they have the been loaded correctly.

\begin{lstlisting}[
    language=Python, 
    caption={Loading MNIST dataset from NPZ archive},
    label={lst:load_mnist},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
X_data = read_mnist_npz("./data/train.npz")[0] # read the training data from the numpy archive
Y_data = read_mnist_npz("./data/train.npz")[1] # read the rraining target data from the numpy archive
X_test = read_mnist_npz("./data/test.npz")[0] # read the test data
Y_test = read_mnist_npz("./data/test.npz")[1] # read the test target data
print(f"Dimensions of our feature matrix: {X_data.shape}")
print(f"Dimensions of our target matrix: {Y_data.shape}")
print(f"Dimensions of our test feature matrix: {X_test.shape}")
print(f"Dimensions of our test target matrix: {Y_test.shape}")
\end{lstlisting}

The output is as follows, confirming that the training feature matrix has 60 000 samples, 784 features, and the target matrix has 60 000 samples and 10 classes (digits 0-9). This differs to the test data, which has 10 000 samples in both the feature and target matrices.
\begin{lstlisting}[style=terminal, caption={Code execution output}]
Dimensions of our training feature matrix: (60000, 784)
Dimensions of our training target matrix: (60000, 10)
Dimensions of our test feature matrix: (10000, 784)
Dimensions of our test target matrix: (10000, 10)
\end{lstlisting}

\section{Question 2: Linear Regression Model Implementation}

\subsection{Normal Equations}
The normal equations are used to determine a closed form solution for the weights of the linear regression model. 
This is implented in the class function \texttt{train\_normaleqs(self, X, Y)}, which takes the feature matrix \texttt{X} and the target matrix \texttt{Y} as input. 
The weights are calculated using the formula:
\begin{equation}
    \mathbf{w}_{\text{opt}} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{y}
\end{equation}
It is important to note that a bias term is added to the feature matrix \texttt{X} so as to account for the intercept in the linear regression model. This is done by concatenating a column of ones to the original feature matrix.
\begin{lstlisting}[
    language=Python, 
    caption={Computing optimal weights using normal equations},
    label={lst:optimal_weights},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
X_bias = self.add_bias(X) # add bias term to input data
weights_opt = np.linalg.inv(X_bias.T @ X_bias) @ X_bias.T @ Y # compute the optimal weights using the normal equations
self.weights = weights_opt # set the model weights to the optimal weights
\end{lstlisting}

\subsection{Forward Pass}
The output of the model $\mathbf{\hat{y}}$ is determined by performing a forward pass through the model. This is implemented in the class function 
\texttt{forward(self, X)} which takes the feature matrix \texttt{X} as input. The output is calculated using the formula:
\begin{equation}
    \mathbf{\hat{y}} = \mathbf{X} \mathbf{w}_{\text{opt}}
\end{equation}
\begin{lstlisting}[
    language=Python, 
    caption={Performing a forward pass through the linear regression model},
    label={lst:forward_pass},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
X_bias = self.add_bias(X) # Add the bias column to input data
y_out = X_bias @ self.weights # Forward pass step
return y_out
\end{lstlisting}

\subsection{Create and train the model}
An instance of the linear regression model is created by calling the class constructor. Both the dimensionality of the model and targets is passed into the constructor. 
The model is then trained (optimal weights are determined) using the normal equations. 

\begin{lstlisting}[
    language=Python, 
    caption={Creating and training the linear regression model using normal equations},
    label={lst:train_model},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
lr = LinearRegression(X_data.shape[1], Y_data.shape[1]) # create the model with D = 784 and K = 10
lr.train_normaleqs(X_data,Y_data) # train the model using the normal equations
\end{lstlisting}

\section{Question 3: Accuracy of the model}
\subsection{Evaluation of model trained using the normal equations}
Using the trained model, predictions are made on the test data. The accuracy of the model is then evaluated using the function \texttt{accuracy(Y,Y\_hat)}, which determines 
the ratio of correct predictions to total predictions. This performance metric is then printed to the terminal. 
\begin{lstlisting}[
    language=Python, 
    caption={Compute accuracy of trained linear regression model},
    label={lst:accuracy},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
y_hat = lr.forward(X_test) # compute the model predictions for the test data
print(f"Trained model accuracy: {accuracy(Y_test, y_hat)}") # Compare ground truth labels to model predictions
\end{lstlisting}
\begin{lstlisting}[style=terminal, caption={Code execution output}]
Trained model accuracy: 0.8604
\end{lstlisting}

\subsection{Baseline comparison}
The model is initialised with random weights so as to assess the performance of an untrained model. Predictions are made on the test data using the untrained model, and the accuracy is computed and printed to the terminal.
\begin{lstlisting}[
    language=Python, 
    caption={Compute accuracy of untrained linear regression model},
    label={lst:optimal_weights},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
lr_untrained = LinearRegression(X_data.shape[1], Y_data.shape[1])
y_hat_untrained = lr_untrained.forward(X_test) # compute the model predictions for the test data
print(f"Untrained model accuracy: {accuracy(Y_test, y_hat_untrained)}")
\end{lstlisting}
\begin{lstlisting}[style=terminal, caption={Code execution output}]
Untrained model accuracy: 0.1083
\end{lstlisting}
Considering that the model is performing a 10 class classification task, the accuracy of the untrained model is consistent with intuition
- random guessing for a 10 class classification problem should result in success 1/10 times. The trained model performs significantly better 
than the untrained model, demonstrating that it has successfully learned to classify the digits in the MNIST dataset.

\section{Question 4: Error of the model}
The error of the untrained model is computed using the function \texttt{error(y, y\_hat)}, which calculates the squared error between the ground truth labels and the model predictions.
\begin{lstlisting}[
    language=Python, 
    caption={Compute error of untrained linear regression model},
    label={lst:untrained_model},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
def error(y, y_hat):
    """
    Calculate the squared error.

    Arguments:
    ----------
        y: np.array of shape (N,K)
            - One-hot encoded targets
        y_hat: np.array of shape (N,K)
            - Digit probabilities/likelihoods for each example in y
    """
    return np.sum((y - y_hat)**2)

rand_pred_error = error(Y_test, y_hat_untrained)
print(f"Random model error: {rand_pred_error}")
\end{lstlisting}
\begin{lstlisting}[style=terminal, caption={Code execution output}]
Random model error: 14307.17285177822
\end{lstlisting}

\section{Question 5: Iterative Gradient Descent}
Another method for training the linear regression model is implemented - iterative gradient descent optimisation. 
This is computed by iteratively computing new weights using the formulae:
\begin{equation}
    \mathbf{w}_{n+1} = \mathbf{w}_n + \Delta \mathbf{w}
\end{equation}
\begin{equation}
    \Delta \mathbf{w} = -\alpha \frac{2}{N} \mathbf{X}^\top (\mathbf{X} \mathbf{w}_n - \mathbf{y})
\end{equation}
This functionality is implemented in the class function \texttt{train(self,X,Y,epochs=200,learning\_rate=0.01)} with the following code:
\begin{lstlisting}[
    language=Python, 
    caption={Gradient descent optimisation for training the linear regression model},
    label={lst:optimal_weights},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
n = X.shape[0] # number of samples
X_bias = self.add_bias(X) # add bias term to input data
        
for epoch in range(epochs):
  y_hat = self.forward(X) # compute the model predictions for the current weights
  change_weights = (2/n) * X_bias.T @ (y_hat - Y) # compute the gradient of the weights
  self.weights -= learning_rate * change_weights # update the weights by moving in the direction of the negative gradient
\end{lstlisting}

\subsection{Create and train model}
An instance of the linear regression model is created and trained using the gradient descent method using the default learning rate and number of training epochs. 
Model predictions are made on both the training and test sets.

\begin{lstlisting}[
    language=Python, 
    caption={Create instance of model},
    label={lst:optimal_weights},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
lr_grad_desc = LinearRegression(X_data.shape[1], Y_data.shape[1])
lr_grad_desc.train(X_data, Y_data)
y_pred_train = lr_grad_desc.forward(X_data)
y_pred_test = lr_grad_desc.forward(X_test)
\end{lstlisting}

\subsection{Model error}
The error of the trained model on both the training and test sets are then computed and printed to the terminal.
\begin{lstlisting}[
    language=Python, 
    caption={Evaluate error of model},
    label={lst:optimal_weights},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
print(f"Error on training set: {error(Y_data, y_pred_train)}")
print(f"Error on test set: {error(Y_test, y_pred_test)}")
\end{lstlisting}

\begin{lstlisting}[style=terminal, caption={Code execution output}]
Error on training set: 27498.859998085863
Error on test set: 4503.876210534427
\end{lstlisting}

Note that the error on the test set is significantly lower than the error on the training set. This is liekly due to the sample sizes of the respective datasets - a larger dataset will result in a higher total squared error value since no normalisation is performed in the error function.

\subsection{Train model with alternate hyperparameters}
\subsubsection{Increase learning rate}
Increasing the learning rate of our gradient descent algorithm results in low accuracy. This is indicative of the algorithm diverging. When the weight 
updates are too large, the algorithm will 'overshoot' the global minimum of the loss function. Instead of converging, the weights oscillate with increasing magnitude, resulting in a model that performs worse than random guessing. This is reflected in the accuracy of the model.
\begin{lstlisting}[
    language=Python, 
    caption={Train model with high learning rate},
    label={lst:highLearning},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
high_lr = LinearRegression(X_data.shape[1], Y_data.shape[1])
high_lr.train(X_data, Y_data, learning_rate=0.1)
y_pred_high_lr = high_lr.forward(X_data)
print(f"Trained model using gradient descent with high learning rate accuracy: {accuracy(Y_data, y_pred_high_lr)}")
\end{lstlisting}
\begin{lstlisting}[style=terminal, caption={Code execution output}]
Trained model using gradient descent with high learning rate accuracy: 0.09751666666666667
\end{lstlisting}

\subsubsection{Decrease learning rate}
The significant increase in accuracy demonstrates that a smaller step size allows the gradient descent algorithm to descend the cost function toward a local or global minimum.
However, given that the accuracy is 63\% - significantly lower than the accuracy obtained using the normal equations - it is likely that the model requires more iterations to reach its optimal state. 
Thus, the weight updates are small and precise but require more iterations to converge.
\begin{lstlisting}[
    language=Python, 
    caption={Train model with low learning rate},
    label={lst:lowLearning},
    breaklines=true,
    basicstyle=\ttfamily\small,
    frame=single
]
low_lr = LinearRegression(X_data.shape[1], Y_data.shape[1])
low_lr.train(X_data, Y_data, learning_rate=0.001)
y_pred_low_lr = low_lr.forward(X_data)
print(f"Trained model using gradient descent with low learning rate accuracy: {accuracy(Y_data, y_pred_low_lr)}")
\end{lstlisting}
\begin{lstlisting}[style=terminal, caption={Code execution output}]
Trained model using gradient descent with low learning rate accuracy: 0.6345166666666666
\end{lstlisting}

\end{document}
